---
title: "Introduction to jBmi with NHANES"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to jBmi with NHANES}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r, echo = FALSE, message = FALSE}
library(jBmi)
library(dplyr)
library(birk)
```

## Data

First, let's start with the dataset, `jBmi::nhanes`.

```{r}
nhanes
```

## Units

Weight and height are in kilograms and meters, respectively. Age is in months. The calculations in this package are in those units.
If your data are in other units, I recommend using `conv_unit()` in the `birk` package.

```{r}
nhanes %>% mutate(`age (yr)` = conv_unit(age, from = "mon", to = "yr"),
                  `height (in)` = conv_unit(height, from = "m", to = "inch"),
                  `weight (lb)` = conv_unit(weight, from = "kg", to = "lbs"))
```

## Calculating BMI

Sticking with metric, we can first calculate BMI. Note that the formula for BMI is simply:

$$\frac{\text{weight (kg)}}{\text{height (m)}^2} = \frac{\text{weight (lb)}}{\text{height (in)}^2} \cdot 703$$

However, the function `calcBMI()` takes an optional `age` argument.
This is because BMI is not meaningful for children under 24 months old.
If age is supplied, then weight-for-age will be calculated for ages less than 24 months.

```{r}
nhanes %>% mutate(type = ifelse(age < 24, "weight-for-age", "bmi"),
                  bmi = calcBMI(height, weight, age))
```

Of course, BMI can still be calculated without considering age.

```{r}
nhanes %>% filter(age >= 24) %>%
  mutate(bmi = calcBMI(height, weight))
```

BMI is essentially the same regardless of whether it was calculated using the metric or US units formula.
Any differences can be chalked up to the approximation in the conversion factor, 703.

```{r}
nhanes %>% filter(age >= 24) %>%
  mutate(`bmi (metric)` = calcBMI(height, weight),
         `height (ft)` = conv_unit(height, from = "m", to = "inch"),
         `weight (lb)` = conv_unit(weight, from = "kg", to = "lbs"),
         `bmi (US)` =  `weight (lb)` / `height (ft)`^2 * 703,
         Difference = `bmi (metric)` - `bmi (US)`) %>%
  select(contains("bmi"), Difference)
```

## Calculating Z-Score

Z-Score is an intermediate statistic to percentile, which is more commonly used.
Z-Scores are calculated using parameters provided by the CDC. They're available for various metrics but
the common protocol is to use weight-for-age for children under 24 months, and BMI for children 24 to 240 months (2 to 20 years).
CDC Z-Scores, growth charts, and percentiles are not made available for adults. 

Z-Scores depend on the measurement (either BMI or weight-for-age), age, and sex.

```{r}
nhanes %>% filter(age <= 240) %>%
  mutate(bmi = calcBMI(height, weight, age),
         zscore = calcZScore(age, sex, bmi),
         percentile = pnorm(zscore))
```

## Detection of Outliers and Data to Review

The `isValid()` function validates the data by flagging data that does meets certain criteria. This function should be used to aid in *reviewing* data, not blindly *removing* data. `isValid()` returns a logical vector and answers the question, "Can I assume this data is valid?" The function returns `FALSE` when:

- Age is 24 months or greater and BMI is less than 6 or greater than 100
- Age is between 24 and 240 months inclusively, and Z-Score is less than -4 or greater than 5
- Age is less than 24 months and Z-Score is less than -6 or greater than 5

These criteria are dependent on age, BMI, and Z-Score, therefore `isValid()` is a function of `age`, `bmi`, and `zscore`.

```{r}
nhanes %>% mutate(bmi = calcBMI(height, weight, age),
                  zscore = calcZScore(age, sex, bmi),
                  percentile = pnorm(zscore)) %>%
  filter(!isValid(age, bmi, zscore))
```

Data of 6 babies were flagged for review. Based on the percentiles, all 6 were heavy for their age.
Additional variables and notes from the NHANES data set should be pulled.
Are the recumbent lengths and head circumferences consistent with the weights?
Are those 5 newborns or was there an error recording the age?
Actually, I suspect that these weights were erroneously recorded in lbs instead of kg.

```{r}
nhanes %>% mutate(bmi = calcBMI(height, weight, age),
                  zscore = calcZScore(age, sex, bmi),
                  percentile = pnorm(zscore)) %>%
  filter(!isValid(age, bmi, zscore)) %>%
  mutate(weight = conv_unit(weight, from = "lbs", to = "kg"),
         bmi = calcBMI(height, weight, age),
         zscore = calcZScore(age, sex, bmi),
         percentile = pnorm(zscore))
```